---
title: "Annand Final Project Code"
author: "Joseph Annand"
date: "2023-12-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries


```{r}
library(ISLR2)
library(MASS)
library(glmnet)
library(pls)
library(leaps)
library(boot)
```


## Load liver data set


```{r}
# Import liver disorder data

liver <- read.table("bupa.data", sep=",", encoding = "UTF-8")
colnames(liver) <- c("mcv", "alkphos", "sgpt", "sgot", "gammagt", "drinks", "selector")

liver <- as.data.frame(liver)
```


## Correlation matrix and multiple linear regression model


```{r}
# Check correlations between predictors and the number of drinks
liver_cor <- cor(liver[,-7], use="complete.obs")
print(liver_cor[6, ])

# Multiple linear regression with all predictors
lm.liver <- lm(drinks ~ ., data = liver[,-7])
summary(lm.liver)
```


The predictors with the strongest correlation to number of drinks are gammagt, mcv, and sgot, respectively. Comparatively, sgpt and alkphos are not strongly correlated to number of drinks.

The F-statistic in our linear model is well above 1, so there is a relationship between the predictors and the response. The statistically significant predictors are mcv and gammagt, while sgot has a p-value just above the significance level.


## Load wine data sets


```{r}
# Import wine quality data

white.quality <- read.csv("winequality-white.csv", sep=";", na.strings = "?", stringsAsFactors = T)
View(white.quality)

red.quality <- read.csv("winequality-red.csv", sep=";", na.strings = "?", stringsAsFactors = T)
View(red.quality)
```



```{r}
white_cor <- cor(white.quality, use="complete.obs")
print(white_cor[12, ])
```


Highest correlations to the quality of the wine can be seen in the alcohol and density predictor variables, followed by chlorides, volatile acidity, and total sulfur dioxide.


```{r}
lm.white <- lm(quality ~ ., data = white.quality)
summary(lm.white)
```


The F-statistic for the multiple linear regression with all predictors is well above 1, meaning our regression fit is statistically significant. The predictors that are statistically significant and have a p-value less than 0.05 are residual sugar, free sulfur dioxide, density, pH, sulphates, and alcohol.