---
title: "Annand Module 04 Lab 01"
author: "Joseph Annand"
date: "2023-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Libraries

```{r}
library(ISLR2)
library(MASS)
library(boot)
```


## Question 6

### Import Data

```{r}
default.data <- Default

set.seed(1)
train.default <- sample(nrow(default.data), nrow(default.data) / 2)
```


### Part A

```{r}
glm.default <- glm(default ~ income + balance, data = default.data,
                   subset = train.default, family = binomial)

summary(glm.default)
```


### Part B

```{r}
boot.fn <- function(data, index)
  + coef(glm(default ~ income + balance, data = data,
             subset = index, family = binomial))

boot.fn(default.data, train.default)
```


### Part C

```{r}
set.seed(9)
boot(default.data, boot.fn, 1000)
```


## Question 9

### Import Data

```{r}
boston.data <- Boston
```


### Part A

```{r}
u_hat <- mean(boston.data$medv)
u_hat
```


### Part B

```{r}
se_hat <- sd(boston.data$medv) / sqrt(nrow(boston.data))
se_hat
```


### Part C

```{r}
u_medv.fn <- function(data, index) {
  z <- data$medv[index]
  sum(z) / length(z)
}

# Bootstrap calcuation for part b
set.seed(8)
boot.b <- boot(boston.data, u_medv.fn, R = 1000)
boot.b
```

The standard errors form part b and the boostrap are very similar with the bootstrap estimate being only about 0.007 units less.


### Part D

```{r}
# Bootstrap from part c estimates mean = 22.53281 and std err = 0.3990518
low_ci <- 22.53281 - 2*0.3990518
upper_ci <- 22.53281 + 2*0.3990518

# Use boot.ci() function to estimate confidence intervals
boot.ci(boot.b, conf = 0.95, type="all")

# Use t-test to estimate confidence intervals
t.test(boston.data$medv)
```

Manually estimating confidence interval and using the t-test yield practically identical results. The estimate from the boot.ci() function is slightly different but more or less the same.


### Part E

```{r}
med_hat <- median(boston.data$medv)
med_hat
```


### Part F

```{r}
med_medv.fn <- function(data, index) {
  median(data$medv[index])
}

set.seed(10)
boot.f <- boot(boston.data, med_medv.fn, 1000)
boot.f
```

The bootstrap approach estimated the same value for the median as in part E. The standard error for the median is slightly lower than that of bootstrap estimation for mean in part C.


### Part G

```{r}
# Get tenth percentile of medv
u_10 <- quantile(boston.data$medv, probs = 0.1, na.rm = FALSE)
u_10
```


### Part H

```{r}
ten_medv.fn <- function(data, index) {
  quantile(data$medv, probs = 0.1, na.rm = FALSE)
}

set.seed(2)
boot.h <- boot(boston.data, ten_medv.fn, 1000)
boot.h
```

The bootstrap approach matches the estimate for the tenth percentile of medv from part G exactly.